---
title: "Reproducible Research Assignment 1"
author: "Nikolaos Lamprou"
date: "August 10, 2014"
output: html_document
---

### Loading and preprocessing the data

Load the data and have a quick look at it.

```{r}
activity<-read.csv("activity.csv",sep=",",header=TRUE)
```

Have a quick look at the data

```{r}
head(activity)
```

Change the data types to assist with the analysis
- Change date from string to POSIXlt object

```{r}
activity$date<-as.POSIXlt(activity$date,format='%Y-%m-%d')
```


### What is mean total number of steps taken per day?

Ignoring missing values make a histogram of the total number of steps taken each day

```{r}
daymonthMatrix<-with(activity,tapply(steps,list(format(date,format='%m'),date$mday),sum))
daymonthDF<-as.data.frame(as.table(daymonthMatrix))
colnames(daymonthDF)<-c("Month","Day", "TotalSteps")
daymonthDF<-daymonthDF[order(daymonthDF$Month,daymonthDF$Day),]
daymonthDF<-daymonthDF[!is.na(daymonthDF$TotalSteps),]
hist(daymonthDF$TotalSteps,col="red",breaks=6,xlab="Total Number of Steps",main="Histogram depicting total number of steps per day")
```


Mean of the total number of steps taken per day
```{r}
mean(daymonthDF$TotalSteps)
```

Median of the total number of steps taken per day
```{r}
median(daymonthDF$TotalSteps)
```

### What is the average daily activity pattern?

Τime series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
avgintervalMatrix<-with(activity,tapply(steps,list(interval),mean,na.rm=TRUE))
avgintervalDF<-as.data.frame(as.table(avgintervalMatrix))
colnames(avgintervalDF)<-c("interval","avg_steps")

avgintervalDF$interval<-as.integer(levels(avgintervalDF$interval))
plot(avgintervalDF$interval,avgintervalDF$avg_steps,ylab="Average number of steps",xlab="5 min intevals",col="red",type="l")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
avgintervalDF[avgintervalDF$avg_steps==max(avgintervalDF$avg_steps),]
```

### Inputing missing values

Total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
nrow(activity[is.na(activity$steps),])
```

Strategy for filling in all of the missing values in the dataset: Use the mean for that 5-minute interval.

```{r}
activity_complete<- merge(activity,avgintervalDF,by="interval")
activity_complete<-activity_complete[order(activity_complete$date,activity_complete$interval),]
for(i in 1:nrow(activity_complete)){
    if(is.na(activity_complete[i,"steps"])){
        activity_complete[i,"steps"]<-activity_complete[i,"avg_steps"]
        }
    }
activity_complete<-activity_complete[,c("steps","date","interval")]
 head(activity_complete)
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity_new<-activity_complete[,c("steps","date","interval")]
head(activity_new)
```

Remake histogram of the total number of steps taken each day.

```{r}
daymonthMatrix<-with(activity_new,tapply(steps,list(format(date,format='%m'),date$mday),sum))
daymonthDF<-as.data.frame(as.table(daymonthMatrix))
colnames(daymonthDF)<-c("Month","Day", "TotalSteps")
daymonthDF<-daymonthDF[order(daymonthDF$Month,daymonthDF$Day),]
daymonthDF<-daymonthDF[!is.na(daymonthDF$TotalSteps),]
hist(daymonthDF$TotalSteps,col="red",breaks=6,xlab="Total Number of Steps",main="Histogram depicting total number of steps/day for complete dataset")
```

Mean of the total number of steps taken per day
```{r}
mean(daymonthDF$TotalSteps)
```

Median of the total number of steps taken per day
```{r}
median(daymonthDF$TotalSteps)
```

The mean and median of the total number of steps remains rougly the same. From the two histograms it is apparent that the distribution is similar but the frequency of observed events is higher.

###Are there differences in activity patterns between weekdays and weekends?


```{r}
weekdayMatrix<-with(activity_new,tapply(steps,list(format(date,format='%a'),interval),mean))
weekdayDF<-as.data.frame(as.table(weekdayMatrix))
colnames(weekdayDF)<-c("day","interval","steps")
head(weekdayDF)
```

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
dtype<-c()
weekdayDF$day<-as.character(weekdayDF$day)
for(j in 1:nrow(weekdayDF)){
    if(weekdayDF[j,"day"]=="Sat"){
        dtype<-c(dtype,"weekend")
    }
    else if(weekdayDF[j,"day"]=="Sun"){
        dtype<-c(dtype,"weekend")
    }
    else{
        dtype<-c(dtype,"weekday")
    }
}
weekdayDF["day_type"]<-dtype
head(weekdayDF,7)
```



Calculate average steps depending on type of day

```{r}
daytypeMatrix<-with(weekdayDF,tapply(steps,list(day_type,interval),mean))
daytypeDF<-as.data.frame(as.table(daytypeMatrix))
colnames(daytypeDF)<-c("day_type","interval","steps")
head(daytypeDF)
```


Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r, fig.height=6,fig.width=12}
library(ggplot2)

ggplot(daytypeDF, aes(x=interval, y=steps, group=day_type)) +
    labs(title="Comparison Between Steps per Interval \n Between Weekends and Weekdays")+
    scale_x_discrete(breaks = seq(from = 0, to = 2355, by = 200))+
    facet_grid(.~ day_type) +
geom_line()+
    ylab("steps")
```

